Ethereum: Un registre de transactions généraliste, sécurisé et décentralisé\\ {\smaller \textbf{révision Homestead}}

### 2016-10-09 Paper.tex
ligne 308

The world state (\textit{state}), is a mapping between addresses (160-bit identifiers) and account states (a data structure serialised as RLP, see Appendix \ref{app:rlp}). Though not stored on the blockchain, it is assumed that the implementation will maintain this mapping in a modified Merkle Patricia tree (\textit{trie}, see Appendix \ref{app:trie}). The trie requires a simple database backend that maintains a mapping of bytearrays to bytearrays; we name this underlying database the state database. This has a number of benefits; firstly the root node of this structure is cryptographically dependent on all internal data and as such its hash can be used as a secure identity for the entire system state. Secondly, being an immutable data structure, it allows any previous state (whose root hash is known) to be recalled by simply altering the root hash accordingly. Since we store all such root hashes in the blockchain, we are able to trivially revert to old states.

The account state comprises the following four fields:

\begin{description}
\item[nonce] A scalar value equal to the number of transactions sent from this address or, in the case of accounts with associated code, the number of contract-creations made by this account. For account of address $a$ in state $\boldsymbol{\sigma}$, this would be formally denoted $\boldsymbol{\sigma}[a]_n$.
\item[balance] A scalar value equal to the number of Wei owned by this address. Formally denoted $\boldsymbol{\sigma}[a]_b$.
\item[storageRoot] A 256-bit hash of the root node of a Merkle Patricia tree that encodes the storage contents of the account (a mapping between 256-bit integer values), encoded into the trie as a mapping from the Keccak 256-bit hash of the  256-bit integer keys to the RLP-encoded 256-bit integer values. The hash is formally denoted $\boldsymbol{\sigma}[a]_s$.
\item[codeHash] The hash of the EVM code of this account---this is the code that gets executed should this address receive a message call; it is immutable and thus, unlike all other fields, cannot be changed after construction. All such code fragments are contained in the state database under their corresponding hashes for later retrieval. This hash is formally denoted $\boldsymbol{\sigma}[a]_c$, and thus the code may be denoted as $\mathbf{b}$, given that $\texttt{\small KEC}(\mathbf{b}) = \boldsymbol{\sigma}[a]_c$.
\end{description}

Since I typically wish to refer not to the trie's root hash but to the underlying set of key/value pairs stored within, I define a convenient equivalence:
\begin{equation}
\texttt{\small TRIE}\big(L_I^*(\boldsymbol{\sigma}[a]_\mathbf{s})\big) \equiv \boldsymbol{\sigma}[a]_s
\end{equation}

The collapse function for the set of key/value pairs in the trie, $L_I^*$, is defined as the element-wise transformation of the base function $L_I$, given as:
\begin{equation}
L_I\big( (k, v) \big) \equiv \big(\texttt{\small KEC}(k), \texttt{\small RLP}(v)\big)
\end{equation}

where:
\begin{equation}
k \in \mathbb{B}_{32} \quad \wedge \quad v \in \mathbb{P}
\end{equation}

It shall be understood that $\boldsymbol{\sigma}[a]_\mathbf{s}$ is not a `physical' member of the account and does not contribute to its later serialisation.

If the \textbf{codeHash} field is the Keccak-256 hash of the empty string, i.e. $\boldsymbol{\sigma}[a]_c = \texttt{\small KEC}\big(()\big)$, then the node represents a simple account, sometimes referred to as a ``non-contract'' account.

Thus we may define a world-state collapse function $L_S$:
\begin{equation}
L_S(\boldsymbol{\sigma}) \equiv \{ p(a): \boldsymbol{\sigma}[a] \neq \varnothing \}
\end{equation}
where
\begin{equation}
p(a) \equiv  \big(\texttt{\small KEC}(a), \texttt{\small RLP}\big( (\boldsymbol{\sigma}[a]_n, \boldsymbol{\sigma}[a]_b, \boldsymbol{\sigma}[a]_s, \boldsymbol{\sigma}[a]_c) \big) \big)
\end{equation}

This function, $L_S$, is used alongside the trie function to provide a short identity (hash) of the world state. We assume:
\begin{equation}
\forall a: \boldsymbol{\sigma}[a] = \varnothing \; \vee \; (a \in \mathbb{B}_{20} \; \wedge \; v(\boldsymbol{\sigma}[a]))
\end{equation}
where $v$ is the account validity function:
\begin{equation}
\quad v(x) \equiv x_n \in \mathbb{P}_{256} \wedge x_b \in \mathbb{P}_{256} \wedge x_s \in \mathbb{B}_{32} \wedge x_c \in \mathbb{B}_{32}
\end{equation}

@@

A significant block number for compatibility with the public network is the block marking the transition between the {\it Frontier} and {\it Homestead} phases of the platform, which we denote with the symbol \firsthomesteadblock, defined thus
\begin{equation}
\firsthomesteadblock \equiv 1,\! 150,\! 000
\end{equation}
The protocol was upgraded at this block, so this symbol appears in some equations to account for the changes.

\subsection{The Transaction} \label{ch:transaction}

A transaction (formally, $T$) is a single cryptographically-signed instruction constructed by an actor externally to the scope of Ethereum. While is assumed that the ultimate external actor will be human in nature, software tools will be used in its construction and dissemination\footnote{Notably, such `tools' could ultimately become so causally removed from their human-based initiation---or humans may become so causally-neutral---that there could be a point at which they rightly be considered autonomous agents. \eg contracts may offer bounties to humans for being sent transactions to initiate their execution.}. There are two types of transactions: those which result in message calls and those which result in the creation of new accounts with associated code (known informally as `contract creation'). Both types specify a number of common fields:

\begin{description}
\item[nonce] A scalar value equal to the number of transactions sent by the sender; formally $T_n$.
\item[gasPrice] A scalar value equal to the number of Wei to be paid per unit of \textit{gas} for all computation costs incurred as a result of the execution of this transaction; formally $T_p$.
\item[gasLimit] A scalar value equal to the maximum amount of gas that should be used in executing this transaction. This is paid up-front, before any computation is done and may not be increased later; formally $T_g$.
\item[to] The 160-bit address of the message call's recipient or, for a contract creation transaction, $\varnothing$, used here to denote the only member of $\mathbb{B}_0$ ; formally $T_t$.
\item[value] A scalar value equal to the number of Wei to be transferred to the message call's recipient or, in the case of contract creation, as an endowment to the newly created account; formally $T_v$.
\item[v, r, s] Values corresponding to the signature of the transaction and used to determine the sender of the transaction; formally $T_w$, $T_r$ and $T_s$. This is expanded in Appendix \ref{app:signing}.
\end{description}

Additionally, a contract creation transaction contains:

\begin{description}
\item[init] An unlimited size byte array specifying the EVM-code for the account initialisation procedure, formally $T_\mathbf{i}$.
\end{description}

\textbf{init} is an EVM-code fragment; it returns the \textbf{body}, a second fragment of code that executes each time the account receives a message call (either through a transaction or due to the internal execution of code). \textbf{init} is executed only once at account creation and gets discarded immediately thereafter.

In contrast, a message call transaction contains:

\begin{description}
\item[data] An unlimited size byte array specifying the input data of the message call, formally $T_\mathbf{d}$.
\end{description}

@@

@@

Here, we assume all components are interpreted by the RLP as integer values, with the exception of the arbitrary length byte arrays $T_\mathbf{i}$ and $T_\mathbf{d}$.
\begin{equation}
\begin{array}[t]{lclclc}
T_n \in \mathbb{P}_{256} & \wedge & T_v \in \mathbb{P}_{256} & \wedge & T_p \in \mathbb{P}_{256} & \wedge \\
T_g \in \mathbb{P}_{256} & \wedge & T_w \in \mathbb{P}_5 & \wedge & T_r \in \mathbb{P}_{256} & \wedge \\
T_s \in \mathbb{P}_{256} & \wedge & T_\mathbf{d} \in \mathbb{B} & \wedge & T_\mathbf{i} \in \mathbb{B}
\end{array}
\end{equation}
where
\begin{equation}
\mathbb{P}_n = \{ P: P \in \mathbb{P} \wedge P < 2^n \}
\end{equation}

The address hash $T_\mathbf{t}$ is slightly different: it is either a 20-byte address hash or, in the case of being a contract-creation transaction (and thus formally equal to $\varnothing$), it is the RLP empty byte-series and thus the member of $\mathbb{B}_0$:
\begin{equation}
T_t \in \begin{cases} \mathbb{B}_{20} & \text{if} \quad T_t \neq \varnothing \\
\mathbb{B}_{0} & \text{otherwise}\end{cases}
\end{equation}

>>>>

L'état mondial (\textit{l'état}), est un mapping entre les adresses (identifiants de 160-bit) et l'état des comptes (une structure de données serialisée comme RLP, voir annexe \ref{app:rlp}). Bien que pas stockée sur la blockchain, on suppose que l'implémentation va maintenir ce mapping dans un arbre Merkle Patricia (\textit{trie}, voir annexe \ref{app:trie}). Le (TODO) trie nécessite une base de donnée en arrière plan qui maintient un mapping de tableau de bytes à tableau de bytes; nous nommons cette base de donnée sous jacente la base de donnée d'état. Elle a un grand nombre d'avantages; premièrement le noeud racine de cette structure est cryptographiquement dépendant de toute la donnée interne et son hash peut être utilisé comme une identité sécurisée pour l'état du système entier. Deuxièmement, étant une structure de donnée immutable, elle autorise n'importe quel état précédent (dont le hash racine est connu) à être appelé à nouveau (TODO : vérifier) en altérant simplement le hash racine accordingly (TODO). Comme nous stockons de tel hash racines dans la blockchain, nous sommes capable de revenir aux anciens états de manière triviale.

L'état d'un compte comprend les quatre champs suivants:

\begin{description}
\item[nonce] Une valeur scalaire égale au nombre de transactions envoyées de cette adresse, ou, dans le cas de compte contenant du code, le nombre de création de contrats faits par ce compte. pour les comptes ou les adresses $a$ dans l'état $\boldsymbol{\sigma}$, cela devra être représenté par $\boldsymbol{\sigma}[a]_n$.
\item[balance] Une valeur scalaire égale au nombre de Wei possédés par cette adresse. noté $\boldsymbol{\sigma}[a]_b$.
\item[storageRoot] Un hash de 256-bit du noeud racine d'un arbre merkle Patricia qui encode le stockage des contenus du compte (un mapping entre des valeurs entières de 256 bits), encodés dans le (TODO) tries comme un mapping entre le hash Keccak 256 bits des clés entières de 256 bits et les valeurs entières encodées par RPL de 256 bits. le hash est noté $\boldsymbol{\sigma}[a]_s$.
\item[codeHash] Le hash du code EVM de ce compte---c'est le code qui est exécuté lorsque cette adresse reçoit un appel de message; c'est immutable et ainsi, contrairement à tous les autres champs, ne peut pas être changé après construction. Tous ces fragments de code sont contenus dans la base de donnée d'état sous leur hash correspondant pour les récupérer plus tard. Ce hash est noté $\boldsymbol{\sigma}[a]_c$, et ainsi le code peut petre représenté par $\mathbf{b}$, sachant que $\texttt{\small KEC}(\mathbf{b}) = \boldsymbol{\sigma}[a]_c$.
\end{description}

Je souhaite typiquement faire référence, non pas hash racine du trie (TODO) mais à l'ensemble sous-jacent de paires de clés/valeurs stockée à l'intérieur, je définis une équivalence convenable: 
\begin{equation}
\texttt{\small TRIE}\big(L_I^*(\boldsymbol{\sigma}[a]_\mathbf{s})\big) \equiv \boldsymbol{\sigma}[a]_s
\end{equation}

la fonction d'effondrement (TODO) pour l'ensemble de paires de clé/valeur dans le trie (TODO), $L_I^*$, est définie comme la transformation element-wise (TODO) de la fonction de base $L_I$, définie par:
\begin{equation}
L_I\big( (k, v) \big) \equiv \big(\texttt{\small KEC}(k), \texttt{\small RLP}(v)\big)
\end{equation}

où:
\begin{equation}
k \in \mathbb{B}_{32} \quad \wedge \quad v \in \mathbb{P}
\end{equation}

Il doit être compris que $\boldsymbol{\sigma}[a]_\mathbf{s}$ n'est pas un membre `physique' du compte et ne contribue pas a sa serialisation ultérieure.

Si le champ \textbf{codeHash} est le hash Keccak-256 de la string vide, par exemple $\boldsymbol{\sigma}[a]_c = \texttt{\small KEC}\big(()\big)$, alors le noeud représente un simple compte, parfois appelé comme un compte ``non-contrat''.

Ainsi nous devons définir une fonction d'effondrement de l'état du monde $L_S$:
\begin{equation}
L_S(\boldsymbol{\sigma}) \equiv \{ p(a): \boldsymbol{\sigma}[a] \neq \varnothing \}
\end{equation}
où
\begin{equation}
p(a) \equiv  \big(\texttt{\small KEC}(a), \texttt{\small RLP}\big( (\boldsymbol{\sigma}[a]_n, \boldsymbol{\sigma}[a]_b, \boldsymbol{\sigma}[a]_s, \boldsymbol{\sigma}[a]_c) \big) \big)
\end{equation}

cette fonction , $L_S$, est utilisée tout au long de la fonction de trie (TODO: trie) pour fournir une identité courte (hash) à l'état du monde. Nous supposons:
\begin{equation}
\forall a: \boldsymbol{\sigma}[a] = \varnothing \; \vee \; (a \in \mathbb{B}_{20} \; \wedge \; v(\boldsymbol{\sigma}[a]))
\end{equation}
où $v$ est la fonction de validité de compte:
\begin{equation}
\quad v(x) \equiv x_n \in \mathbb{P}_{256} \wedge x_b \in \mathbb{P}_{256} \wedge x_s \in \mathbb{B}_{32} \wedge x_c \in \mathbb{B}_{32}
\end{equation}

@@

Un numéro de bloc significatif pour la compatibilité avec le réseau public est le bloc marquant la transition entre les phases {\it Frontier} et {\it Homestead} de la plateforme, que nous représentons avec le symbole \firsthomesteadblock, défini ainsi 
\begin{equation}
\firsthomesteadblock \equiv 1,\! 150,\! 000
\end{equation}
Le protocole a été mis à jour à ce bloc, donc ce symbole apparait dans quelques équations de comptes, pour les changements.

\subsection{La Transaction} \label{ch:transaction}

Une transaction ($T$) est une instruction de signature cryptographique unique construite par un acteur externe du spectre d'Ethereum. Alors que nous supposons que l'ultime acteur externe sera humain par nature, les outils logiciels qui seront utilisé pour sa construction et sa propagation\footnote{Particulièrement, de tels `outils' pourraient en fin de compte devenir causalement écartées de leur création humaine---ou les humains peuvent devenir causalement neutres---de telle manière qu'il pourraient se comporter comme des agents autonomes. \eg Les contrats peuvent offrir des primes aux humains pour avoir envoyé des transactions pour initier leur exécution.}. Il y a deux types de transactions: celles qui entraînent des appels de message et celles qui entraînent la création de nouveaux comptes auxquels est associé du code (connu de manière informelle sous le nom de `création de contrat'). Chacun de ces types de transaction spécifient un nombre de champs communs:

\begin{description}
\item[nonce] Une valeur scalaire égale au nombre de transactions émises par l'envoyeur $T_n$.
\item[gasPrice] Une valeur scalaire égale au nombre de Wei à payer par unité de \textit{gas} pour tous les coûts de calcul engagés comme résultat de l'exécution de cette transaction; $T_p$.
\item[gasLimit] Une valeur scalaire égale au montant maximum de gas qui peut re utilisé en exécutant cette transaction. Cela est payé en amont, avant que tout calcul soit fait et ne peut pas être augmenté plus tard; $T_g$.
\item[to] L'adresse de 160 bits du destinataire de l'appel de message ou, pour une transaction de création de contrat , $\varnothing$, utilisé ici pour noter le seul membre de $\mathbb{B}_0$ ; formellement $T_t$.
\item[value] Une valeur scalaire égale au nombre de Wei à transférer au destnataire de l'appel de message ou, dans le cas d'une création de contrat, comme une dotation pour le nouveau compte créé; $T_v$.
\item[v, r, s] Les valeurs correspondantes à la signature de la transaction et utilisées pour déterminer l'emetteur de la transaction; formellement $T_w$, $T_r$ and $T_s$. Cela est détaillé dans l'annexe \ref{app:signing}.
\end{description}

De plus, une transaction de création de contrat contient:

\begin{description}
\item[init] Un tableau d'octets de taille infinie contenant le code EVM pour la procédure d'initialisation du compte, $T_\mathbf{i}$.
\end{description}

\textbf{init} est un fragment de code EVM; il retourne le \textbf{body}, un second fragment de code qui s'exécute à chaque fois que le compte reçoit un appel de message (soit au travers d'une transaction, soit par l'exécution interne de code). \textbf{init} est exécuté une seule fois à la création du compte et reste écarté immédiatement après.

En revanche, une transaction d'appel de message contient:

\begin{description}
\item[data] Un tableau d'octets illimité spécifiant la donnée entrante de l'appel de message, $T_\mathbf{d}$.
\end{description}

@@

L'annexe \ref{app:signing} décrit la fonction, $S$, qui fait correspondre les transactions au destinataire, et s'exécute à travers l'ECDSA de la courbe SECP-256k1, utilisant le hash de la transaction (excepté les trois derniers champs de signature) en tant que datum à signer. Pour le moment nous nous assurons seulement que l'émetteur d'une transaction donnée peut être représenté par $S(T)$.

@@

Ici, nous supposons que tous les composants sont interprétés par le RPL comme des valeurs entières, avec l'exception d'octets de longueur arbitraire $T_\mathbf{i}$ et $T_\mathbf{d}$.

\begin{equation}
\begin{array}[t]{lclclc}
T_n \in \mathbb{P}_{256} & \wedge & T_v \in \mathbb{P}_{256} & \wedge & T_p \in \mathbb{P}_{256} & \wedge \\
T_g \in \mathbb{P}_{256} & \wedge & T_w \in \mathbb{P}_5 & \wedge & T_r \in \mathbb{P}_{256} & \wedge \\
T_s \in \mathbb{P}_{256} & \wedge & T_\mathbf{d} \in \mathbb{B} & \wedge & T_\mathbf{i} \in \mathbb{B}
\end{array}
\end{equation}
où
\begin{equation}
\mathbb{P}_n = \{ P: P \in \mathbb{P} \wedge P < 2^n \}
\end{equation}

Le hash de l'adresse $T_\mathbf{t}$ est légèrement différent: c'est soit le hash d'une adresse de 20 octets, ou, dans le cas d'une création de contrat (et ainsi formellement égal à $\varnothing$), c'est la série RPL d'octets vides et ainsi le membre de $\mathbb{B}_0$:

\begin{equation}
T_t \in \begin{cases} \mathbb{B}_{20} & \text{if} \quad T_t \neq \varnothing \\
\mathbb{B}_{0} & \text{otherwise}\end{cases}
\end{equation}
